Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2019-08-25 Manuscript/manuscript.Rmd
edits

We analyzed publically available replicate Hi-C data sets from the GM12878 and K562 cell lines (Supplementary Table 1). Topologically associating domains (TADs) were subsequently called using the ARROWHEAD pipeline [@durand2016juicer] at 5 data resolutions: 5 kb 10 kb, 25 kb, 50 kb, and 100 kb. After removing TADs that exceeded 2 megabases in length, we obtained a total of 16133, 14824, 8081, 3560, and 919 unique TAD boundaries for the GM12878 cell line, and 9314, 10881, 6923, 3197, and 831 unique TAD boundaries for the K562 cell line across each of the 5 resolutions respectively (Supplementary Figure 1). To account for the variability in boundary location due to varying resolution, we developed a resolution-specific approach toward constructing TAD versus non-TAD boundary regions on the linear genome (See Methods), labeled as a response vector **Y** (Figure 1A). TAD boundary regions were defined as regions on the linear genome where a genomic bin overlapped with a flanked TAD boundary ($Y = 1$), while non-TAD boundary regions were defined as the regions that did not overlap with a flanked TAD boundary ($Y = 0$). A significant number of TAD boundary regions were found to be specific to one of the two cell lines (Supplementary Figure 2). Given the plasticity in boundary location, we opted to build cell line specific models for predicting TAD boundary regions. The results presented are for the GM12878 cell line, with K562 results presented as supplementary.

![**Illustrating the resolution-specific data construction and feature engineering used for downstream modelling.** (A) Unique boundaries are denoted by the blue dots. The yellow regions denote the flanked boundary regions according to the Hi-C data resolution (i.e. 100 kb resolution corresponded to a 100 kb flank for a total boundary region of 200 kb). The linear genome was subsequently binned into resolution-specific intervals. The response vector **Y** was defined as a variable that indicated whether a genomic bin overlapped with a flanked TAD boundary ($Y = 1$) or not ($Y = 0$). (B) Three types of associations between TAD boundaries and genomic annotations including _Overlap Counts_ (OC), _Overlap Percent_ (OP), and ($log_{2}$) _Distance_. The green shapes denote regions defined by a specific functional genomic annotation used for feature engineering. ](Figures/model_construction_and_predictor_types.png)

@@

We then investigated the proportional class imbalance (PI) featured at each resolution. The PI was defined as the ratio of TAD boundary regions ($Y=1$) versus the sum of both TAD and non-TAD boundary regions for each chromosome specific sections of the linear genome and has a range from 0 to 0.5. The PI was found to be invariant to resolution, remaining consistently low, with an average of 0.06, 0.10, 0.14, 0.12, and 0.06 for the 5 kb, 10 kb, 25 kb, 50 kb, and 100 kb resolutions respectively. The K562 cell line was found to have comparatively low proportional imbalances (Supplementary Figure 3). Such low proportions of imbalance gave precedence for the implementation of re-sampling prior to performing classification.

@@

A total of 78 cell type specific functional genomic annotations derived from ChIP-Seq experiments (See Methods) for the GM12878 and K562 cell line were used to build the feature space of downstream predictive models (Supplementary Tables 2-3). We constructed three different predictor types for each functional genomic annotation. Counts of overlaps (**overlap count**, OC) were defined as the total number of peak regions that overlapped within a given genomic bin (Figure 1B). Percents of overlaps (**overlap percent**, OP) were defined as the proportion of a bin overlapped by a peak region (Figure 3B). **Distance** was defined as log2 transformed number of base pairs between between the centers of bins and regions defining genomic annotations (Figure 3C). A log2-transformation of distance was used to account for the skewness present in the distributions of distances <!--(Supplementary Figures 1-2)-->. Each predictor type provides a granular way to draw associations between functional genomic annotations and the spatial structure of the genome.

@@

In order to identify the optimal combination of data resolution, predictor type, and re-sampling technique necessary for TAD boundary prediction, we created an ensemble framework to systematically compare performances of different model inputs (Figure 4). Each system in the framework was assumed to be composed of a random forest classifier constructed from a particular cell line (*L = {GM12878, K562}*). Within each cell line we evaluated 5 data resolutions (*R = {5 kb, 10 kb, 25 kb, 50 kb, 100 kb}*). We then considered 3 separate predictor types as the feature space (*P = {OC, OP, Distance}*). To assess the severity of class imbalance, 4 re-sampling techniques were implemented prior to modelling (*B = {None, RUS, ROS, SMOTE}*). Each combination of the above cell line specific parameters was implemented on chromosome specific sections of the linear genome (omitting chromosome 9). Letting $|X|$ denote the cardinality [??? TBD] of each set, we evaluated *|L|* $\times$ *|R|* $\times$ *|P|* $\times$ *|B|* different ensemble systems, totaling 2520 (*2* $\times$ *5* $\times$ *3* $\times$ *4* $\times$ *21*) separate models. For each combination of inputs, we performed 10-fold cross-validation to tune for the number of features to consider at each node of the random forest classifier. Each optimally tuned model was then validated on the same separate cell line specific testing set for unbiased comparisons. Performances were evaluated using MCCs and aggregated across each chromosome by averaging. <!--We report our main results for the GM12878 cell line and have included results from the K562 cell line in the supplementary materials.-->

>>>>

We analyzed publically available replicate Hi-C data sets from the GM12878 and K562 cell lines (Supplementary Table 1). Topologically associating domains (TADs) were subsequently called using the ARROWHEAD pipeline [@durand2016juicer] at 5 data resolutions: 5 kb 10 kb, 25 kb, 50 kb, and 100 kb. After removing TADs that exceeded 2 megabases in length, we obtained a total of 16133, 14824, 8081, 3560, and 919 unique TAD boundaries for the GM12878 cell line, and 9314, 10881, 6923, 3197, and 831 unique TAD boundaries for the K562 cell line across each of the 5 resolutions respectively (Supplementary Figure 1). To account for the variability in boundary location due to varying resolution, we developed a resolution-specific approach toward constructing TAD versus non-TAD boundary regions on the linear genome (See Methods), labeled as a response vector **Y** (Figure 1A). TAD boundary regions were defined as regions on the linear genome where a genomic bin overlapped with a flanked TAD boundary ($Y = 1$), while non-TAD boundary regions were defined as the regions that did not overlap with a flanked TAD boundary ($Y = 0$). A significant number of TAD boundary regions were found to be specific to one of the two cell lines (Supplementary Figure 2). Given the plasticity in boundary location, we opted to build cell line specific models for predicting TAD boundary regions. The results presented are for the GM12878 cell line, with K562 results presented in the supplementary materials.

![**Illustrating the resolution-specific data construction and feature engineering used for downstream modeling.** (A) Unique boundaries are denoted by the blue dots. The yellow regions denote the flanked boundary regions according to the Hi-C data resolution (i.e. 100 kb resolution corresponded to a 100 kb flank for a total boundary region of 200 kb). The linear genome was subsequently binned into resolution-specific intervals. The response vector **Y** was defined as a genomic bin that either overlapped with a flanked TAD boundary ($Y = 1$) or did not ($Y = 0$). (B) Three types of associations between TAD boundaries and genomic annotations were used to build the feature space including _Overlap Counts_ (OC), _Overlap Percent_ (OP), and ($log_{2}$) _Distance_. The green shapes denote regions defined by a specific functional genomic annotation used for feature engineering. ](Figures/model_construction_and_predictor_types.png)

@@

We then investigated the proportional class imbalance (PI) featured at each resolution. The PI was defined as the ratio of TAD boundary regions ($Y=1$) versus the sum of both TAD and non-TAD boundary regions for each chromosome specific sections of the linear genome and has a range from 0 to 0.5. The PI was found to be invariant to resolution, remaining consistently low, with an average of 0.06, 0.10, 0.14, 0.12, and 0.06 for the 5 kb, 10 kb, 25 kb, 50 kb, and 100 kb resolutions respectively. The K562 cell line was found to have comparatively low proportional imbalances (Supplementary Figure 3). Such low PI gave precedence for the implementation of re-sampling prior to performing classification.

@@

A total of 78 cell type specific functional genomic annotations derived from ChIP-Seq experiments (See Methods) for the GM12878 and K562 cell line were used to build the feature space of downstream predictive models (Supplementary Tables 2-3). The annotations could be broken down into 3 sets including histone modifications (HM), chromatin state segmentations (ChromHMM), and transcription factor binding sites (TFBS). We constructed three different predictor types for each functional genomic annotation (Figure 1B). Counts of overlaps (**overlap count**, OC) were defined as the total number of peak regions that overlapped within a given genomic bin. Percents of overlaps (**overlap percent**, OP) were defined as the proportion of a bin overlapped by a peak region. **Distance** was defined as log2 transformed number of base pairs between between the centers of bins and regions defining genomic annotations. A log2-transformation of distance was used to account for the skewness present in the distributions of distances (Supplementary Figures 4-5).

@@

In order to identify the combination of data resolution, predictor type, and re-sampling technique necessary for optimal cell line specific TAD boundary prediction, we created an ensemble framework to systematically compare performances of different model inputs (Figure 2). Each system in the framework was assumed to be composed of a random forest classifier constructed from a particular cell line (*L = {GM12878, K562}*). Within each cell line we evaluated 5 data resolutions (*R = {5 kb, 10 kb, 25 kb, 50 kb, 100 kb}*). We then considered 3 separate predictor types as the feature space (*P = {OC, OP, Distance}*). To assess the severity of class imbalance, 4 re-sampling techniques were implemented prior to modelling (*B = {None, RUS, ROS, SMOTE}*). Once the model inputs were established a random forest classifier was trained on chromosome specific sections of the genome using 10-fold cross validation. The model was then validated on the same separate chromosome specific test set for unbiased comparisons and evaluated using Matthews Correlation Coefficient (MCC). Letting $|X|$ denote the cardinality of each set, we evaluated *|L|* $\times$ *|R|* $\times$ *|P|* $\times$ *|B|* different ensemble systems, totaling 2520 (*2* $\times$ *5* $\times$ *3* $\times$ *4* $\times$ *21*) separate models. Additional performance metrics are reported in the supplementary materials.
